Legion: Expressing Locality and Independence with Logical Regions

### 2012-07-03 doc/popl2013/langdef.tex
some edits suggested by Alex

\begin{figure*}
\centering{
\framebox{$\typeenv{bv}{bool}$}
\framebox{$\typeenv{iv}{int}$}
\finfrule
{\begin{array}{l}
\typeenvx e_1 : T_1 \\
\typeenvx e_2 : T_2
\end{array}}
{\typeenvx \langle e_1, e_2 \rangle : \langle T_1, T_2 \rangle}
\finfrule{\typeenvx e : \langle T_1,T_2 \rangle}{\typeenvx e\text{.1}\ : T_1}
\finfrule{\typeenvx e : \langle T_1,T_2 \rangle}{\typeenvx e\text{.2}\ : T_2}
\finfrule{\Gamma(id) = T}{\typeenvx id : T}
\framebox{$\typeenvx \text{null }T@r : T@r$}
\framebox{$\typeenvx \text{new }T@r : T@r$}
\finfrule{\typeenvx e : T@(\oton{r})}{\typeenvx \text{isnull}(e) : bool}
\finfrule
{\begin{array}{l}
\typeenvx e : T@(\otok{k}{r'}) \\
\forall i. \exists j, r'_i \leq r_j \in \Omega^* \\
\end{array}}
{\typeenvx upregion(e,\oton{r}) : T@(\oton{r})}
\finfrule
{\begin{array}{l}
\typeenvx e : T@(\otok{k}{r'}) \\
\forall j. \exists i, r_j \leq r'_i \in \Omega^* \\
\end{array}}
{\typeenvx downregion(e,\oton{r}) : T@(\oton{r})}
\finfrule
{\begin{array}{l}
\typeenvx e_1 : T@(\oton{r}) \\
\forall i, reads(r_i) \in \Phi^*\end{array}}
{\typeenvx \text{read}(e_1) : T}
\finfrule
{\begin{array}{l}
\typeenvx e_1 : T@(\oton{r}) \\
\typeenvx e_2 : T \\
\forall i, writes(r_i) \in \Phi^*
\end{array}}
{\typeenvx \text{write}(e_1, e_2) : bool}
\finfrule
{\begin{array}{l}
\Gamma(id) = (T_1, T_2), \emptyset, \emptyset \rightarrow T_1 \\
\typeenvx e_1 : T_1@(\oton{r}) \\
\typeenvx e_2 : T_2 \\
\forall i, reduces(id, r_i) \in \Phi^*
\end{array}}
{\typeenvx \text{reduce}(id, e_1, e_2) : bool}
\framebox{$\typeenvx \text{newcolor }r : \text{coloring}(r)$}
\finfrule{\begin{array}{l}
\typeenvx e_1 : \text{coloring}(r) \\
\typeenvx e_2 : T@r \\
\typeenvx e_3 : int
\end{array}}
{\typeenvx \text{color}(e_1, e_2, e_3) : \text{coloring}(r)}
\finfrule{\begin{array}{l}\typeenvx e_1 : int \\ \typeenvx e_2 : int\end{array}}{\typeenvx e_1 + e_2 : int}
\finfrule{\begin{array}{l}\typeenvx e_1 : int \\ \typeenvx e_2 : int\end{array}}{\typeenvx e_1 < e_2 : bool}
\finfrule{\begin{array}{l}
\typeenvx e_1 : T_1 \\
\Gamma' = \Gamma[id/T_1] \\
\typeenvx[G=\Gamma'] e_2 : T_2
\end{array}}
{\typeenvx : \text{let}\ id : T_1 \text{in}\ e_2 : T_2}
\finfrule{\begin{array}{l}\typeenvx e_1 : bool \\ \typeenvx e_2 : T \\ \typeenvx e_3 : T\end{array}}{\typeenvx \text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3 : T}
\finfrule{
\begin{array}{l}
\Gamma(id) = \forall \otok{k}{r'}.(\oton{T}),\Phi_i, Q \rightarrow T_r \\
\typeenvx e_i : T_i[r_1/r'_1,\ldots,r_k/r'_k] \\
\Phi_i[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Phi^*
\end{array}}
{\typeenvx id[\otok{k}{r}](\oton{e}) : T_r[r_1/r'_1,\ldots,r_k/r'_k]}
\finfrule{
\begin{array}{l}
\typeenvx e_1 : \text{coloring}(r_p) \\
\Omega' = \Omega \wedge \bigwedge_{i \in [1,k]} r_i \leq r_p \wedge \bigwedge_{i \not= j} r_i * r_j \\
\typeenvx[O=\Omega'] e_2 : T \\
r_i \not\in \textit{regions\_of}(\Gamma, T)
\end{array}}
{\typeenvx \text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2 : T}
\finfrule{
\begin{array}{l}
T_1 = \exists \otok{k}{r'}. T_2\text{ where }\Omega_1 \\
\Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Omega^* \\
\typeenvx e_1 : T_2[r_1/r'_1,\ldots,r_k/r'_k]
\end{array}}
{\typeenvx \text{pack}\ e_1\ \text{as}\ T_1[\otok{k}{r}] : T_1}
\finfrule{
\begin{array}{l}
T_1 = \exists \otok{k}{r'}. T_2\text{ where }\Omega_1 \\
\typeenvx e_1 : T_1 \\
\Gamma' = \Gamma[T_2[r_1/r'_1,\ldots,r_k/r'_k] / id] \\
\Omega' = \Omega \wedge \Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \\
\typeenvx[G=\Gamma',O=\Omega'] e_2 : T_3 \\
r_i \not\in \textit{regions\_of}(\Gamma, T_1, T_3)
\end{array}}
{\typeenvx \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 : T_3}
}
\caption{Legion Core Type System}
\label{fig:types}
\end{figure*}

\begin{figure}
\centering{
$\begin{array}{l}
\Omega \subseteq \Omega^* \\
r_i \leq r_j \in \Omega^*  \Rightarrow r_i \leq r_i \in \Omega^*\wedge r_j \leq r_j \in \Omega^* \\
r_i \leq r_j \in \Omega^* \wedge r_j \leq r_k \in \Omega^* \Rightarrow r_i \leq r_k \in \Omega^* \\
r_i \leq r_j \in \Omega^* \wedge r_j * r_k \in \Omega^* \Rightarrow r_i * r_k \in \Omega^* \\
r_i * r_j \in \Omega^* \Rightarrow r_j * r_i \in \Omega^* \\
\\
\Phi \subseteq \Phi^* \\
r_i \leq r_j \in \Omega^* \wedge reads(r_j) \in \Phi^* \Rightarrow reads(r_i) \in \Phi^* \\
r_i \leq r_j \in \Omega^* \wedge writes(r_j) \in \Phi^* \Rightarrow writes(r_i) \in \Phi^* \\
r_i \leq r_j \in \Omega^* \wedge reduces(id, r_j) \in \Phi^* \Rightarrow reduces(id, r_i) \in \Phi^* \\
reads(r) \in \Phi^* \wedge writes(r) \in \Phi^* \wedge \\
\hspace{1cm} \Gamma(id) = \forall \otok{k}{r}. (\oton{T}), E \rightarrow T_r \Rightarrow reduces(id, r) \in \Phi^*
\end{array}$
}
\caption{Privilege and Constraint Closure}
\label{fig:closure}
\end{figure}

@@

The following helper functions are used in some of the rules that deal with memory effects:
\begin{itemize}
\item $apply(S, E, C)$ is used to determine the effect on the store of a sequence of memory operations
\item $mark\_coherence(E, \hat Q)$ marks the memory operations in a subtask's trace to
constrain how the operations in that trace can be interleaved with other sibling tasks' operations
\item $valid\_interleave(S, C, E', \oton{E})$ is used to determine if $E'$ is a valid way to
interleave the memory traces of two or more subexpressions
\end{itemize}
Figure~\ref{fig:opsemfns} gives the definitions of {\em apply} and {\em mark\_coherence}.  For
the {\em valid\_interleave} predicate, we state some of the necessary conditions here, but leave 
the construction of actual valid interleavings until a later section.  Informally, a interleaving
of two or more memory traces must include all the operations from each trace (and no additional
operations), with each trace's operations remaining in order.  A valid interleaving must further
satisfy the requirement that the value read from the store must yield the most recently 
stored/reduced value unless that location is part of the clobber set $C$.
\end{itemize}

@@

\begin{figure*}
\centering{
$
\begin{array}{lll}
apply(S, [], C) & = & S \\
apply(S, E \dplus [ read(l, c, v) ]) & = & apply(S, E, C) \\
apply(S, E \dplus [ write(l, c, v) ]) & = & apply(S, E, C)[v/l] \\
apply(S, E \dplus [ reduce_{id}(l, c, v) ]) & = & apply(S, E, C)[id(S[l], v)/l] \\
mark\_coherence([], \hat Q) & = & [] \\
mark\_coherence(E \dplus [ op(l, c, v) ], \hat Q) & = & mark\_coherence(E, \hat Q) \dplus \\
& & op(l, c', v), \text{ where } \\
& & c' =
\begin{cases}
simult, & \text{if }\exists \rho. l \in \rho \wedge simult(\rho) \in \hat Q \\
atomic, & \text{if }\exists \rho. l \in \rho \wedge atomic(\rho) \in \hat Q \\
excl, & \text{otherwise}
\end{cases} \\
any\_interleave(E', [], \ldots, []) & \rightarrow & true \\
any\_interleave([ op(l, c, v) ] \dplus E', E_1, \ldots, [ op(l, c, v) ] \dplus E_i, \ldots, E_n) & \rightarrow & any\_interleave(E', E_1, \ldots, E_i, \ldots, E_n) \\
valid\_interleave(S, C, E', \oton{E}) & \rightarrow & any\_interleave(E', \oton{E}) \\
valid\_interleave(S, C, E' \dplus [ read(l, c, v) ] \dplus E'', \oton{E}) & \rightarrow & l \in C \vee S'(l) = v, \text{ where } S' = apply(S, E', C)
\end{array}
$}
\caption{Operational Semantics Helper Functions}
\label{fig:opsemfns}
\end{figure*}

@@

S' = apply(S, E_1, C) \\

@@

{\opsenvx \text{isnull}(e) \mapsto false, E}

@@

S' = \text{apply}(S, E, C) \\

@@

S' = \text{apply}(S, E_1, C) \\

@@

S'' = \text{apply}(S', E_2, C) \\

@@

false, & \text{otherwise}.

@@

\opsenvx e_1 \mapsto false, E_1 \\
S' = \text{apply}(S, E_1, C) \\

@@

S_1 = \text{apply}(S, E_1, C) \\

@@

S_n = \text{apply}(S_{n-1}, E_n, C) \\

@@

\text{task }id[\otok{k}{r'}](f_1 : T_1, \ldots, f_n : T_n), \Phi_i, Q : T_r = e_{n+1} \\

@@

L' = \{ (f_1, v_1), \ldots, (f_n, v_n) \} \\
S' = \text{apply}(S, E', C) \\
C' = C \cup \{ l : \exists \rho. atomic(\rho) \in M' \llbracket Q \rrbracket \vee simult(\rho) \in M' \llbracket Q \rrbracket \} \\

@@

E'_{n+1} = mark\_coherence(E_{n+1}, M' \llbracket Q \rrbracket) \\

@@

\rho_i = \{ l : (l, i) \in K \} \\

@@

\rho_i = M[r_i] \\

@@

\begin{lem}
\label{lemma:heapconst:apply}
If $S \sim H$ and $E \sim H$, then $\text{apply}(S, E, C) \sim H$.
\end{lem}

>>>>

\begin{figure*}
\centering{
\framebox{$\typeenv{bv}{bool}$}
\framebox{$\typeenv{iv}{int}$}
\finfrule
{\begin{array}{l}
\typeenvx e_1 : T_1 \\
\typeenvx e_2 : T_2
\end{array}}
{\typeenvx \langle e_1, e_2 \rangle : \langle T_1, T_2 \rangle}
\finfrule{\typeenvx e : \langle T_1,T_2 \rangle}{\typeenvx e\text{.1}\ : T_1}
\finfrule{\typeenvx e : \langle T_1,T_2 \rangle}{\typeenvx e\text{.2}\ : T_2}
\finfrule{\Gamma(id) = T}{\typeenvx id : T}
\framebox{$\typeenvx \text{null }T@r : T@r$}
\framebox{$\typeenvx \text{new }T@r : T@r$}
\finfrule{\typeenvx e : T@(\oton{r})}{\typeenvx \text{isnull}(e) : bool}
\finfrule
{\begin{array}{l}
\typeenvx e : T@(r'_1, \ldots r'_k) \\
\forall i. \exists j, r'_i \leq r_j \in \Omega^* \\
\end{array}}
{\typeenvx upregion(e,\oton{r}) : T@(\oton{r})}
\finfrule
{\begin{array}{l}
\typeenvx e : T@(r'_1, \ldots r'_k) \\
\forall j. \exists i, r_j \leq r'_i \in \Omega^* \\
\end{array}}
{\typeenvx downregion(e,\oton{r}) : T@(\oton{r})}
\finfrule
{\begin{array}{l}
\typeenvx e_1 : T@(\oton{r}) \\
\forall i, reads(r_i) \in \Phi^*\end{array}}
{\typeenvx \text{read}(e_1) : T}
\finfrule
{\begin{array}{l}
\typeenvx e_1 : T@(\oton{r}) \\
\typeenvx e_2 : T \\
\forall i, writes(r_i) \in \Phi^*
\end{array}}
{\typeenvx \text{write}(e_1, e_2) : bool}
\finfrule
{\begin{array}{l}
\Gamma(id) = (T_1, T_2), \emptyset, \emptyset \rightarrow T_1 \\
\typeenvx e_1 : T_1@(\oton{r}) \\
\typeenvx e_2 : T_2 \\
\forall i, reduces(id, r_i) \in \Phi^*
\end{array}}
{\typeenvx \text{reduce}(id, e_1, e_2) : bool}
\framebox{$\typeenvx \text{newcolor }r : \text{coloring}(r)$}
\finfrule{\begin{array}{l}
\typeenvx e_1 : \text{coloring}(r) \\
\typeenvx e_2 : T@r \\
\typeenvx e_3 : int
\end{array}}
{\typeenvx \text{color}(e_1, e_2, e_3) : \text{coloring}(r)}
\finfrule{\begin{array}{l}\typeenvx e_1 : int \\ \typeenvx e_2 : int\end{array}}{\typeenvx e_1 + e_2 : int}
\finfrule{\begin{array}{l}\typeenvx e_1 : int \\ \typeenvx e_2 : int\end{array}}{\typeenvx e_1 < e_2 : bool}
\finfrule{\begin{array}{l}
\typeenvx e_1 : T_1 \\
\typeenvx[G={\Gamma[id/T_1]}] e_2 : T_2
\end{array}}
{\typeenvx : \text{let}\ id : T_1 \text{in}\ e_2 : T_2}
\finfrule{\begin{array}{l}\typeenvx e_1 : bool \\ \typeenvx e_2 : T \\ \typeenvx e_3 : T\end{array}}{\typeenvx \text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3 : T}
\finfrule{
\begin{array}{l}
\Gamma(id) = \forall r'_1, \ldots r'_k.(\oton{T}),\Phi', Q' \rightarrow T_r \\
\typeenvx e_i : T_i[r_1/r'_1,\ldots,r_k/r'_k] \\
\Phi'[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Phi^*
\end{array}}
{\typeenvx id[\otok{k}{r}](\oton{e}) : T_r[r_1/r'_1,\ldots,r_k/r'_k]}
\finfrule{
\begin{array}{l}
\typeenvx e_1 : \text{coloring}(r_p) \\
\Omega' = \Omega \wedge \bigwedge_{i \in [1,k]} r_i \leq r_p \wedge \bigwedge_{1 \leq i < j \leq k} r_i * r_j \\
\typeenvx[O=\Omega'] e_2 : T \\
\{ \oton{r} \} \cap \textit{regions\_of}(\Gamma, T) = \emptyset
\end{array}}
{\typeenvx \text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2 : T}
\finfrule{
\begin{array}{l}
T_1 = \exists r'_1, \ldots r'_n.\ T_2\text{ where }\Omega_1 \\
\Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Omega^* \\
\typeenvx e_1 : T_2[r_1/r'_1,\ldots,r_k/r'_k]
\end{array}}
{\typeenvx \text{pack}\ e_1\ \text{as}\ T_1[\otok{k}{r}] : T_1}
\finfrule{
\begin{array}{l}
T_1 = \exists r'_1, \ldots, r'_n.\ T_2\text{ where }\Omega_1 \\
\typeenvx e_1 : T_1 \\
\Gamma' = \Gamma[T_2[r_1/r'_1,\ldots,r_k/r'_k] / id] \\
\Omega' = \Omega \cup \Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \\
\typeenvx[G=\Gamma',O=\Omega'] e_2 : T_3 \\
\{ \oton{r} \} \cap \textit{regions\_of}(\Gamma, T_1, T_3) = \emptyset
\end{array}}
{\typeenvx \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 : T_3}
}
\caption{Legion Core Type System}
\label{fig:types}
\end{figure*}

\begin{figure}
\centering{
$\begin{array}{l}
\Omega \subseteq \Omega^* \\
r_i \leq r_j \in \Omega^*  \Rightarrow r_i \leq r_i \in \Omega^*\wedge r_j \leq r_j \in \Omega^* \\
r_i \leq r_j \in \Omega^* \wedge r_j \leq r_k \in \Omega^* \Rightarrow r_i \leq r_k \in \Omega^* \\
r_i \leq r_j \in \Omega^* \wedge r_j * r_k \in \Omega^* \Rightarrow r_i * r_k \in \Omega^* \\
r_i * r_j \in \Omega^* \Rightarrow r_j * r_i \in \Omega^* \\
\\
\Phi \subseteq \Phi^* \\
r_i \leq r_j \in \Omega^* \wedge reads(r_j) \in \Phi^* \Rightarrow reads(r_i) \in \Phi^* \\
r_i \leq r_j \in \Omega^* \wedge writes(r_j) \in \Phi^* \Rightarrow writes(r_i) \in \Phi^* \\
r_i \leq r_j \in \Omega^* \wedge reduces(id, r_j) \in \Phi^* \Rightarrow reduces(id, r_i) \in \Phi^* \\
reads(r) \in \Phi^* \wedge writes(r) \in \Phi^* \wedge \\
\hspace{1cm} \Gamma(id) = \forall \otok{k}{r}. (\oton{T}), \Phi', Q' \rightarrow T_r \Rightarrow reduces(id, r) \in \Phi^*
\end{array}$
}
\caption{Privilege and Constraint Closure}
\label{fig:closure}
\end{figure}

@@

The following helper functions are used in some of the rules that deal with memory effects:
\begin{itemize}
\item $apply(S, E)$ is used to determine the effect on the store of a sequence of memory operations
\item $mark\_coherence(E, \hat Q)$ marks the memory operations in a subtask's trace to
constrain how the operations in that trace can be interleaved with other sibling tasks' operations
\item $valid\_interleave(S, C, E', \oton{E})$ is used to determine if $E'$ is a valid way to
interleave the memory traces of two or more subexpressions
\end{itemize}
Figure~\ref{fig:opsemfns} gives the definitions of {\em apply} and {\em mark\_coherence}.  For
the {\em valid\_interleave} predicate, we state some of the necessary conditions here, but leave 
the construction of actual valid interleavings until a later section.  Informally, a interleaving
of two or more memory traces must include all the operations from each trace (and no additional
operations), with each trace's operations remaining in order.  A valid interleaving must further
satisfy the requirement that the value read from the store must yield the most recently 
stored/reduced value unless that location is part of the clobber set $C$.
\end{itemize}

@@

\begin{figure*}
\centering{
$
\begin{array}{lll}
apply(S, []) & = & S \\
apply(S, E \dplus [ read(l, c, v) ]) & = & apply(S, E) \\
apply(S, E \dplus [ write(l, c, v) ]) & = & apply(S, E)[v/l] \\
apply(S, E \dplus [ reduce_{id}(l, c, v) ]) & = & S'[id(S'(l), v)/l], \text{ where } S' = apply(S, E) \\
mark\_coherence([], \hat Q) & = & [] \\
mark\_coherence(E \dplus [ op(l, c, v) ], \hat Q) & = & mark\_coherence(E, \hat Q) \dplus \\
& & op(l, c', v), \text{ where } \\
& & c' =
\begin{cases}
simult, & \text{if }\exists \rho. l \in \rho \wedge simult(\rho) \in \hat Q \\
atomic, & \text{if }\exists \rho. l \in \rho \wedge atomic(\rho) \in \hat Q \\
excl, & \text{otherwise}
\end{cases} \\
any\_interleave(E', [], \ldots, []) & \rightarrow & true \\
any\_interleave([ op(l, c, v) ] \dplus E', E_1, \ldots, [ op(l, c, v) ] \dplus E_i, \ldots, E_n) & \rightarrow & any\_interleave(E', E_1, \ldots, E_i, \ldots, E_n) \\
valid\_interleave(S, C, E', \oton{E}) & \rightarrow & any\_interleave(E', \oton{E}) \\
valid\_interleave(S, C, E' \dplus [ read(l, c, v) ] \dplus E'', \oton{E}) & \rightarrow & l \in C \vee S'(l) = v, \text{ where } S' = apply(S, E')
\end{array}
$}
\caption{Operational Semantics Helper Functions}
\label{fig:opsemfns}
\end{figure*}

@@

S' = apply(S, E_1) \\

@@

{\opsenvx \text{isnull}(e) \mapsto \textit{false}, E}

@@

S' = \text{apply}(S, E) \\

@@

S' = \text{apply}(S, E_1) \\

@@

S'' = \text{apply}(S', E_2) \\

@@

\textit{false}, & \text{otherwise}.

@@

\opsenvx e_1 \mapsto \textit{false}, E_1 \\
S' = \text{apply}(S, E_1) \\

@@

S_1 = \text{apply}(S, E_1) \\

@@

S_n = \text{apply}(S_{n-1}, E_n) \\

@@

\text{task }id[\otok{k}{r'}](a_1 : T_1, \ldots, a_n : T_n), \Phi', Q' : T_r = e_{n+1} \\

@@

L' = \{ (a_1, v_1), \ldots, (a_n, v_n) \} \\
S' = \text{apply}(S, E') \\
C' = C \cup \{ l : \exists \rho. atomic(\rho) \in M' \llbracket Q' \rrbracket \vee simult(\rho) \in M' \llbracket Q' \rrbracket \} \\

@@

E'_{n+1} = mark\_coherence(E_{n+1}, M' \llbracket Q' \rrbracket) \\

@@

\rho_i = \{ l : (l, i) \in K \}, \text{ for } 1 \leq i \leq k \\

@@

\rho_i = M[r_i], \text{ for } 1 \leq i \leq k \\

@@

\begin{lem}
\label{lemma:heapconst:apply}
If $S \sim H$ and $E \sim H$, then $\text{apply}(S, E) \sim H$.
\end{lem}
